---
title: "AF validation"
output: html_notebook
---

Load libraries
```{r}
library(Metrics)
library(tidyverse)
library(patchwork)
library(viridis)
library(parallel)
```

# Standard samtools markdup
Load data
```{r}
# Inds data
inds305 <- read.table("/data/genetics_tmp/mapping_tests/results_NT/duplicate_filtering_AF_validation/AF_tables/ACORN_VCF_Individual_Pool305_AlleleFreqs.frq", header=TRUE, 
                   row.names = NULL, na.strings = "-nan")

colnames(inds305) <- c("chrom", "pos", "alleles", "read_depth", "ref_ind", "alt_ind")
inds305 <- inds305 %>% 
  unite(site, chrom, pos, sep="_")

## Filter for coverage
inds305 <- inds305 %>% 
  filter(read_depth >= 40) %>% 
  select(site, alt_ind)

inds312 <- read.table("/data/genetics_tmp/mapping_tests/results_NT/duplicate_filtering_AF_validation/AF_tables/ACORN_VCF_Individual_Pool312_AlleleFreqs.frq", sep="\t", header=TRUE, 
                   row.names = NULL, na.strings = "-nan")

colnames(inds312) <- c("chrom", "pos", "alleles", "read_depth", "ref_ind", "alt_ind")
inds312 <- inds312 %>% 
  unite(site, chrom, pos, sep="_")

## Filter for coverage
inds312 <- inds312 %>% 
  filter(read_depth >= 40) %>% 
  select(site, alt_ind)

# Pool data
pooldata_markdup <- read.table("/data/genetics_tmp/mapping_tests/results_NT/duplicate_filtering_AF_validation/AF_tables/validationIndelFilteredSNPs_Biallelic_markdup_miss0.1_ADP100_AlleleFrequencyTable.txt", sep = "\t", header=TRUE)
pool305_markdup <- pooldata_markdup %>% 
                   select(chrom_pos, X305_markdup)
colnames(pool305_markdup) <- c("site", "alt_pool")

pool312_markdup <- pooldata_markdup %>% 
                   select(chrom_pos, X312_markdup)
colnames(pool312_markdup) <- c("site", "alt_pool")

# Make consensus data set
consensus305_markdup <- na.omit(inner_join(inds305, pool305_markdup))
consensus312_markdup <- na.omit(inner_join(inds312, pool312_markdup))

poplist <- list(consensus305_markdup, consensus312_markdup)
```

Statistics
```{r}
error_stats <- function(ind_freq, pool_freq){
    # Fit a linear regression model
    model <- lm(ind_freq ~ pool_freq)
    
    # Extract R-squared value from the model summary
    Rsquared<- paste0("R squared = ",summary(model)$r.squared)
    
    # rmse
    rmse <- paste0("RMSE = ",rmse(ind_freq, pool_freq))
    
    # mae
    mae <- paste0("MAE = ",mae(ind_freq, pool_freq))
    
    # mape (Mean Absolute Percentage Error)
    mape <- function(y_true, y_pred) {
      non_zero_indices <- which(y_true != 0)
      if (length(non_zero_indices) > 0) {
        mean(abs((y_true[non_zero_indices] - y_pred[non_zero_indices]) / y_true[non_zero_indices])) * 100
      } else {
        NA
      }
    }
    mape_res <- paste0("MAPE = ",mape(ind_freq, pool_freq))
  out <- paste("\n", Rsquared, rmse, mae, mape_res, sep="\n")
  return(out)
}

# Apply the error_stats function to each element of poplist using parallel processing
results_markdup <- mclapply(poplist, function(x) {
  error_stats(x$alt_ind, x$alt_pool)
}, mc.cores = detectCores() - 1)

# print the results
cat(results_markdup[[1]], results_markdup[[2]])
```

Plots
```{r fig.width=7, fig.height=14}
hist2d <- function(dataset, alt_ind=alt_ind, alt_pool=alt_pool){
#Create the 2D histogram plot:
  p <- ggplot(dataset, aes(x=alt_ind, y=alt_pool)) + 
       geom_bin2d(bins=41) + 
       scale_fill_viridis(trans='log10') +
       theme_minimal() + 
       labs(x="Ind-Seq ALT frequency",
            y="Pool-Seq ALT frequency",
            fill="Log10 Density") + 
       theme(legend.position = c(1.2, 1.15))
  
  #Create marginal histograms:
  px <- ggplot(dataset, aes(x=alt_ind)) + geom_histogram(bins=41) + theme_void()
  py <- ggplot(dataset, aes(x=alt_pool)) + geom_histogram(bins=41) + theme_void() +
                coord_flip()
  
  #Arrange the plots together:
  px + plot_spacer() + p + py +
    plot_layout(
      ncol = 2, nrow = 2,
      widths = c(4,1), heights = c(1,4)
    )
}

histALTfreq <- function(dataset, alt_ind=alt_ind, alt_pool=alt_pool){
  ggplot(dataset) +
  geom_histogram(aes(x=alt_ind, after_stat(density)), bins = 41,
                 fill = "lightblue", color = "black") +
  geom_density(aes(x=alt_pool), color="darkred", size=1) +
  geom_vline(xintercept = .05, color="black", linetype="dashed") +
  theme_minimal() +  
  labs(x="ALT frequency", y="Density",
       title=paste())
}

# res_hist2d <- mclapply(poplist, function(x){
#   hist2d(dataset=x)
# }, mc.cores = detectCores() - 1)

res_ALTfreq_markdup <- mclapply(poplist, function(x){
  histALTfreq(dataset=x)
}, mc.cores = detectCores() - 1)

# res_hist2d[[1]]
# res_hist2d[[2]]

res_ALTfreq_markdup[[1]] / res_ALTfreq_markdup[[2]] + plot_annotation(tag_levels = 'A')
```

Barplot of AF
```{r, eval=FALSE}
# consensus_tidy <- pivot_longer(consensus, names_to="seq_type", 
#                                values_to = "freq", !site)
# 
# ggplot(consensus_tidy, aes(x=freq, fill=seq_type))+
#   geom_histogram(position = "dodge", bins = 41) +
#   theme(legend.position = c(.9, .9))
```

# Deduplication by identifying duplicate sequences with fastp
Load data
```{r}
# Pool data
pooldata_dedup <- read.table("/data/genetics_tmp/mapping_tests/results_NT/duplicate_filtering_AF_validation/AF_tables/validationIndelFilteredSNPs_Biallelic_dedup_NOmarkdup_miss0.1_ADP100_AlleleFrequencyTable.txt", sep = "\t", header=TRUE)
pool305_dedup <- pooldata_dedup %>% 
                   select(chrom_pos, X305_dedup_NOmarkdup)
colnames(pool305_dedup) <- c("site", "alt_pool")

pool312_dedup <- pooldata_dedup %>% 
                   select(chrom_pos, X312_dedup_NOmarkdup)
colnames(pool312_dedup) <- c("site", "alt_pool")

# Make consensus data set
consensus305_dedup <- na.omit(inner_join(inds305, pool305_dedup))
consensus312_dedup <- na.omit(inner_join(inds312, pool312_dedup))

poplist <- list(consensus305_dedup, consensus312_dedup)
```

Statistics
```{r}
# Apply the error_stats function to each element of poplist using parallel processing
results_dedup <- mclapply(poplist, function(x) {
  error_stats(x$alt_ind, x$alt_pool)
}, mc.cores = detectCores() - 1)

# print the results
cat(results_dedup[[1]], results_dedup[[2]])
```

Plots
```{r fig.width=7, fig.height=14}
# res_hist2d <- mclapply(poplist, function(x){
#   hist2d(dataset=x)
# }, mc.cores = detectCores() - 1)

res_ALTfreq_dedup <- mclapply(poplist, function(x){
  histALTfreq(dataset=x)
}, mc.cores = detectCores() - 1)

# res_hist2d[[1]]
# res_hist2d[[2]]

res_ALTfreq_dedup[[1]] / res_ALTfreq_dedup[[2]] + plot_annotation(tag_levels = 'A')
```

# Deduplication by identifying duplicate sequences with fastp and with the alignment-based method implemented in samtools
Load data
```{r}
# Pool data
pooldata_dedup_markdup <- read.table("/data/genetics_tmp/mapping_tests/results_NT/duplicate_filtering_AF_validation/AF_tables/validationIndelFilteredSNPs_Biallelic_dedup_markdup_miss0.1_ADP100_AlleleFrequencyTable.txt", sep = "\t", header=TRUE)
pool305_dedup_markdup <- pooldata_dedup_markdup %>% 
                   select(chrom_pos, X305_dedup_markdup)
colnames(pool305_dedup_markdup) <- c("site", "alt_pool")

pool312_dedup_markdup <- pooldata_dedup_markdup %>% 
                   select(chrom_pos, X312_dedup_markdup)
colnames(pool312_dedup_markdup) <- c("site", "alt_pool")

# Make consensus data set
consensus305_dedup_markdup <- na.omit(inner_join(inds305, pool305_dedup_markdup))
consensus312_dedup_markdup <- na.omit(inner_join(inds312, pool312_dedup_markdup))

poplist <- list(consensus305_dedup_markdup, consensus312_dedup_markdup)
```

Statistics
```{r}
# Apply the error_stats function to each element of poplist using parallel processing
results_dedup_markdup <- mclapply(poplist, function(x) {
  error_stats(x$alt_ind, x$alt_pool)
}, mc.cores = detectCores() - 1)

# print the results
cat(results_dedup_markdup[[1]], results_dedup_markdup[[2]])
```

Plots
```{r fig.width=7, fig.height=14}
# res_hist2d <- mclapply(poplist, function(x){
#   hist2d(dataset=x)
# }, mc.cores = detectCores() - 1)

res_ALTfreq_dedup_markdup <- mclapply(poplist, function(x){
  histALTfreq(dataset=x)
}, mc.cores = detectCores() - 1)

# res_hist2d[[1]]
# res_hist2d[[2]]

res_ALTfreq_dedup_markdup[[1]] / res_ALTfreq_dedup_markdup[[2]] + plot_annotation(tag_levels = 'A')
```

Save data for later referencing
```{r}
#save.image(file = "/data/genetics_tmp/mapping_tests/results_NT/AF_validation.Rdata")
```

