---
title: "pcadapt"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Palettes
```{r}
cb_paired=c("#B1DF8A","#33A02C","#A6CEE3","#1F78B4","#FB9A99","#E31A1C",
            "#CAB2D6","#6A3D9A","#FDBF6F","#FF7F00","grey50","#B15928",
            "black")

xlstat=c("#7a81ff", "#00fe0a", "#7030a0", "#0070c0", "#00b050", "#03f2fc", "#f905ab",
         "#92d050", "#945200", "#ff0000", "#262626", "#c00000", "#ffc000")

wong=c("#000000","#E69F00","#56B4E9","#009E73","#CC79A7","#F0E442","#006699",
         "#D55E00")
```

## Libraries
```{r}
library(pcadapt)
library(qvalue)
library(ggplot2)
library(dplyr)
library(stringr)
```

## Load previous results
```{r}
load("../results/popgen/poolfstat.RData")
```

# Analysis
keep only SNPs found in Chromosomes, NOT scaffolds!
```{r}
#get chromosome info
chrom <- filtered@snp.info
str(chrom)
dim(chrom)

# find rows of chromosome SNPS
chrom <- dplyr::filter(chrom, str_detect(Chromosome, "Qrob_Chr"))

# filter pcadapt input file
pcadapt_input <- ref_freq[ ,1:nrow(chrom)]
```


```{r}
filename <- read.pcadapt(pcadapt_input, type = "pool")
```

```{r}
res <- pcadapt(filename)
#The same as res <- pcadapt(filename,K=2)
summary(res)
```

Scree plot will not be available with only two pops
```{r}
plot(res,option="screeplot")
```

list of outliers
```{r}
padj <- p.adjust(res$pvalues, method = "BH")
alpha <- 0.01
outliers <- which(padj < alpha)
length(outliers)
```

Make Manhattan plot by chromosome
NA p values arise because of MAF filter
MAF filter works differently in poolfstat and pcadapt
In poolfstat MAF filter is global (as if the data consisted of one pop only)
PCAdapt MAF filter is applied per population (pool) basis
```{r fig.width=10, fig.height= 5}
pca_df <- as.data.frame(cbind(chrom, padj))
colnames(pca_df) <- c("chr", "position", "ref", "alt", "pvalues")
pca_df$chr <- gsub(pattern = "Qrob_Chr0", replacement = "", pca_df$chr)
pca_df$chr <- gsub(pattern = "Qrob_Chr", replacement = "", pca_df$chr)
pca_df$chr <- as.factor(reorder(pca_df$chr, sort(as.integer(pca_df$chr))))
str(pca_df)

manh_p <- ggplot(data = pca_df, aes(x = position, 
                                    y = -log10(pvalues),
                                    color = reorder(chr,
                                                    sort(as.numeric(chr))))) +
  facet_wrap(~chr, scales = "free_x", nrow=1, strip.position = "bottom") +
  geom_point(alpha = 1, size = 0.8) +
  geom_abline(intercept = -log10(sort(padj)[length(outliers)]), 
              slope = 0, linetype = "dashed") +
  scale_color_manual(values = rep(wong[-6], 3)) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.spacing = unit(0, "lines")) +
  xlab("Chromosome") + ylab("-log10(p-value)")

manh_p + 
    theme(panel.grid.major.x = element_line(colour = NA,
      linetype = "blank"),
      panel.grid.minor = element_line(linetype = "blank"),
      plot.background = element_rect(fill = "white", color = "white"))

ggsave(filename = "Qrob_pcadapt-p-values-by-chr.png",
       path = "../results/figs", dpi=300,
       width = 10, height = 5)
```

```{r fig.width=10, fig.height=24}
manh_p2 <- ggplot(data = pca_df, aes(x = position / 10^6, 
                                    y = -log10(pvalues),
                                    color = chr)) +
  facet_wrap(~chr, scales = "free", ncol=1, strip.position = "top") +
  geom_point(alpha = 1, size = 0.8) +
  geom_abline(intercept = -log10(sort(padj)[length(outliers)]), 
              slope = 0, linetype = "dashed") +
  scale_color_manual(values = rep(wong[-6], 4)) +
  theme_bw() +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(colour = NA, linetype = "blank"),
        panel.grid.minor = element_line(linetype = "blank"),
        panel.spacing = unit(0, "lines")) +
  xlab("Position (Mb)") + ylab("-log10(p-value)") +
  scale_x_continuous(labels = scales::comma)

manh_p2 + 
    theme(plot.background = element_rect(fill = "white", color = "white"))

ggsave(filename = "Qrob_pcadapt-p-values-by-chr2.png",
       path = "../results/figs", dpi=300,
       width = 10, height = 24)
```


